Enhancing Forecast Reconciliation: A Study of Alternative Covariance Estimators

### 2025-07-22 literature/paper_draft.qmd
paste the cov ests section from research proposal

## The NOVELIST Estimator

$$
g(\hat{W}_1) = \hat{W}^{shr}_{1, thr} = \lambda_{\delta} \hat{W}_{1, \delta} + (1 - \lambda_{\delta}) \hat{W}_1
$$

is the NOVELIST shrinkage estimator, proposed by Huang & Fryzlewicz (2019). By convenient setting, we rewrite as sample correlation:

$$
\hat{R}^{shr}_{1, thr} = \lambda_{\delta} \hat{R}_{1,\delta} + (1 - \lambda_{\delta}) \hat{R}_1,
$$

$\hat{R}_{1, \delta}$ is a thresholded correlation matrix, in which thresholding is applied only to each off-diagonal element. This approach will shrink the sample correlation matrix $\hat{R}_1$ towards its thresholded version. There are various choices for the thresholding function, in this work, we use the soft-thresholding operator, defined as:

$$
\hat{r}_{1,ij}^\delta = \operatorname{sign}(\hat{r}_{1,ij}) \, (|\hat{r}_{1,ij}| - \delta)_+.
$$

After calculated the NOVELIST correlation matrix, we can re-obtain the covariance matrix $\hat{W}^{shr}_{1, thr} = \hat{D}_1^{1/2} \hat{R}^{shr}_{1, thr} \; \hat{D}_1^{1/2}$, where $\hat{D}_1 = diag(\hat{W}_1)$ is the diagonal matrix and the elements are given by the variances of the 1-step-ahead base forecast errors.

For a given threshold $\delta$, the optimal shrinkage intensity parameter $\lambda(\delta)$ can be estimated as:

$$
\hat{\lambda}(\delta) = \frac{
  \sum_{i \neq j} \widehat{Var}(\hat{r}_{1,ij}) \; I(|\hat{r}_{1,ij}| \leq \delta)
} {
  \sum_{i \neq j} (\hat{r}_{1,ij} - \hat{r}_{1,ij}^\delta)^2
}
$$

The formula derived using Ledoit-Wolf's lemma (Ledoit and Wolf, 2003) by Huang & Fryzlewicz. For the threshold parameter $\delta$, we use a cross-validation to select the optimal value

-   ALGORITHM WALKTHROUGH

**Principal-Component Adjustment**

When a factor structure is present, the procedure is:

<!-- 1. *Standardize Forecast Errors:* -->

<!-- $$ -->

<!-- \hat{z}_{t|t-1} = D_1^{-1/2} \hat{e}_{t|t-1}. -->

<!-- $$ -->

<!-- 2. *Construct the Data Matrix:* -->

<!-- $$ -->

<!-- \hat{Z} = [\hat{z}_{1|0}, \hat{z}_{2|1}, \dots, \hat{z}_{T|T-1}] -->

<!-- $$ -->

<!-- 3. *Extract Principal Components:* -->

<!-- Compute the top K eigenvectors to form: -->

<!-- $$ -->

<!-- \hat{B} = T^{-1} \hat{Z} \hat{V}_K, -->

<!-- $$ -->

<!-- where $\hat{V}_K$ contains the K largest eigenvectors. -->

<!-- 4. *Residual Component:* -->

<!-- $$ -->

<!-- \hat{Z}_{\text{rem}} = \hat{Z} - \hat{B}\hat{V}_K^\top. -->

<!-- $$ -->

<!-- Then compute the residual correlation: -->

<!-- $$ -->

<!-- \tilde{R}_1 = T^{-1\prime} (\hat{Z} - \hat{B}\hat{V}_K^\top)(\hat{Z} - \hat{B}\hat{V}_K^\top)^\top. -->

<!-- $$ -->

<!-- 5. *Apply NOVELIST to Residuals:* -->

<!-- $$ -->

<!-- \hat{R}_{K,\text{shr}}^{1,\text{thresh}} = \lambda \tilde{R}_\delta^1 + (1 - \lambda) \tilde{R}_1. -->

<!-- $$ -->

<!-- 6. *Reconstruct Full Estimator:* -->

<!-- $$ -->

<!-- \hat{R}_{1,K} = \hat{B}\hat{B}^\top + \hat{R}_{K,\text{shr}}^{1,\text{thresh}}. -->

<!-- $$ -->

<!-- **Shrinkage Intensity \(\lambda\):**   -->

<!-- Estimated as -->

<!-- $$ -->

<!-- \hat{\lambda} = \max\Big(0, \min\Big(1, \frac{\sum_{i \neq j} \operatorname{Var}(\hat{r}_{1,ij})}{\sum_{i \neq j} \hat{r}_{1,ij}^2}\Big)\Big). -->

<!-- $$ -->

<!-- **Threshold \(\delta\):**   -->

<!-- Selected via grid search or cross-validation. -->

## Alternative Covariance Estimators

Brief overview of other high-dimensional estimators used in the literature for benchmarking.

Others:

-   Principal Orthogonal complEment Thresholding (POET) by Fan et al. (2013)

    A Low-rank + Sparse method.\
    It decompose the covariance matrix into a prominent principle components part (Low-rank) and a orthogonal complement part $R_K$. Then apply thresholding to $R_K$.\
    This is similar to the NOVELIST estimator with PC-adjusted, but the difference is that POET apply thresholding to $R_K$, not a NOVELIST function.

>>>>

## NOVELIST Estimator

The NOVELIST (NOVEL Integration of the Sample and Thresholded Covariance) estimator, proposed by @Huang2019-ua, is currently the main focus of this research project. It introduces a way to control the target matrix's sparsity, retaining strong correlations while discarding weak, noisy effects. NOVELIST offers more flexibility than the shrinkage estimator, which is useful when we believe that only a few variables are truly correlated. However, it does not guarantee to be positive definite.

The method is based on the idea of soft-thresholding the sample covariance matrix, then performing shrinkage towards this thresholded version. This introduces an extra parameter, the threshold $\delta$, which is used to control the amount of soft-thresholding. The NOVELIST estimator is given by:

$$
\hat{\boldsymbol{W}}^{N}_{1} = \lambda_{\delta} \hat{\boldsymbol{W}}_{1, \delta} + (1 - \lambda_{\delta}) \hat{\boldsymbol{W}}_1,
$$ {#eq-novelist-cov}

where $\hat{\boldsymbol{W}}_{1, \delta}$ is the thresholded version of $\hat{\boldsymbol{W}}_{1}$. By convenient setting, we can rewrite it in terms of correlation:

$$
\hat{\boldsymbol{R}}^{N}_{1} = \lambda_{\delta} \hat{\boldsymbol{R}}_{1,\delta} + (1 - \lambda_{\delta}) \hat{\boldsymbol{R}}_1,
$$ {#eq-novelist-cor}

In this setting, $\hat{\boldsymbol{R}}_{1,\delta}$ is the thresholded correlation matrix, where each element is regularised by:

$$
\hat{r}_{1,ij}^\delta = \text{sign}(\hat{r}_{1,ij}) \, 
\text{max}(|\hat{r}_{1,ij}| - \delta, \; 0),
$$ {#eq-soft-thr}

where $\delta \in [0,1]$ is the threshold parameter. For a given threshold $\delta$, @Huang2019-ua derived an analytical expression for the optimal shrinkage intensity parameter $\lambda(\delta)$ using Ledoit-Wolf's lemma [@Ledoit2003-qv], following similar logic to @Schafer2005-yw. It can be computed as:

<!-- - \textcolor{red}{@Huang2019-ua mentioned Ledoit-Wolf's lemma} -->

$$
\hat{\lambda}(\delta) = \frac{
  \sum_{i \neq j} \widehat{Var}(\hat{r}_{1,ij}) \; \boldsymbol{1}(|\hat{r}_{1,ij}| \leq \delta)
} {
  \sum_{i \neq j} (\hat{r}_{1,ij} - \hat{r}_{1,ij}^\delta)^2
},
$$ {#eq-lambda-thr}

where $\boldsymbol{1}(.)$ is the indicator function.

On the other hand, the optimal threshold $\delta^*$ does not have a closed-form solution, and is typically obtained by executing a rolling-window cross-validation procedure. The idea is to find the threshold $\hat{\delta^*}$, with the corresponding $\hat{\lambda}^*$ and $\hat{\boldsymbol{R}}^{N}_{1}(\hat{\delta^*}, \hat{\lambda}^*)$, that minimises the average out-of-sample reconciled forecast errors. The formal algorithm is given in the @sec-novelist_cv Appendix. Although it is not required to fit forecasting models multiple times, the cross-validation procedure is still computationally expensive as it computes the NOVELIST estimator and perform reconciliation for each threshold value.

Note that when $\delta \in \bigl[ \text{max}_{i \neq j}|\hat{r}_{1,ij}|, \; 1 \bigr]$, the NOVELIST estimator collapses to the shrinkage estimator, and when $\delta = 0$, it becomes the sample covariance matrix.

## POET

The POET (Principal Orthogonal complEment Thresholding) estimator, proposed by @Fan2013-jz, is another "sparse" + "non-sparse" covariance estimator. It takes the latent factors directly into its construction, and is appealing when there are common drivers in the time series within the hierarchy, as we saw in the Australian tourism example.

The POET method starts by decomposing the correlation matrix $\hat{\boldsymbol{R}}_1$ into a prominent principle components part (low-rank) and a orthogonal complement part $\hat{\boldsymbol{R}}_{1, K}$ (the correlation matrix after removing the first $K$ principal components). Then it applies thresholding to $\hat{\boldsymbol{R}}_{1, K}$. The POET estimator is given by:

$$
\hat{\boldsymbol{R}}^{K}_{1} = \sum_{k = 1}^K \hat{\gamma}_k \hat{\boldsymbol{\xi}}_k \hat{\boldsymbol{\xi}}_k' + T(\hat{\boldsymbol{R}}_{1, K})
$$

where $\hat{\gamma}_k$ and $\hat{\boldsymbol{\xi}}_k$ are the $k$th largest eigenvalue and the corresponding eigenvector of the sample covariance matrix, respectively, and $T(.)$ is the thresholding function, which can be either soft-thresholding, hard-thresholding, or others.

## PC-adjusted NOVELIST

This approach is best of both worlds, leveraging the strengths of both NOVELIST and POET. The PC-adjusted (Principal-Component-adjusted) NOVELIST overcomes the shortcomings of the current shrinkage estimator, taking prominent PCs into account while also offers extra flexibility. The idea is to apply the NOVELIST estimator to the orthogonal complement part $\hat{\boldsymbol{R}}_{1, K}$, and then add the principal components part back. The PC-adjusted NOVELIST estimator is formulated as:

$$
\hat{\boldsymbol{R}}^{N, K}_{1} = \sum_{k = 1}^K \hat{\gamma}_k \hat{\boldsymbol{\xi}}_k \hat{\boldsymbol{\xi}}_k' + \hat{\boldsymbol{R}}^{N}_{1, K}
\; ,
$$

where $\hat{\boldsymbol{R}}^{N}_{1, K}$ is the NOVELIST estimator applied to the orthogonal complement part $\hat{\boldsymbol{R}}_{1, K}$. Similar to the NOVELIST estimator, $\hat{\boldsymbol{R}}^{N, K}_{1}$ is not guaranteed to be positive definite.

Methods to ensure positive definiteness of the NOVELIST estimator (and its PC-adjusted variant) will be explored and studied in the project. @Huang2019-ua proposed to diagonalise the NOVELIST estimator and replace any eigenvalues that fall under a certain small positive threshold by the value of that threshold. Alternatively, we can implement the algorithm of @Higham2002-te that computes the nearest positive definite matrix to a given matrix.
